A New Score for Adaptive Tests\\ in Bayesian and Credal Networks

### 2021-05-16 paper.tex

\usepackage{algorithm,algorithmic,amsfonts,bm,hyperref,booktabs,tikz,pgf,pgfplots,amsmath,soul,subcaption,url,graphicx,amsmath,booktabs,booktabs,nicefrac,bm,colortbl,tikz,pgfplots,enumerate,xargs}

@@

\captionsetup{compatibility=false}

@@

\title{A New Score for Adaptive Tests\\in Bayesian and Credal Networks}

@@

Francesca Mangili \and\\

@@

\institute{Istituto Dalle Molle di Studi sull'Intelligenza Artificiale, Lugano, Switzerland\\

@@

A test is \emph{adaptive} when its sequence and number of questions is dynamically tuned on the basis of the estimated skills of the taker. Graphical models, such as Bayesian networks, are used for adaptive tests as they allow to model the uncertainty about the questions and the skills in an explainable fashion, especially when coping with multiple skills. A better elicitation of the uncertainty in the question/skills relations can be achieved by interval probabilities. This turns the model into a \emph{credal} network, thus making the inferential complexity of the queries required to select questions more demanding. This is especially the case for the information theoretic quantities used as \emph{scores} to drive the adaptive mechanism. We present an alternative family of scores, based on the mode of the posterior probabilities and hence easier to explain. This makes considerably simpler the evaluation in the credal case, without significantly affecting the quality of the adaptive process. Numerical tests on synthetic and real-world data are used to support this claim.
%12 pages
\keywords{Computer adaptive tests \and Information theory \and Credal networks \and Bayesian networks}

@@

A test or an exam can be naturally intended as a measurement process, with the questions acting as sensors measuring the skills of the test taker in a particular discipline. Such measurement is typically imperfect with the skills acting as latent variables whose actual values cannot be revealed in a perfectly reliable way. The role of the questions, whose answers are regarded instead as manifest variables, is to reduce the uncertainty about the latent skills. Following this perspective, probabilistic modelling is an obvious framework to describe tests. Consider for instance the example in Figure \ref{fig:minicat}, where a small Bayesian network is used to describe the probability of the test taker having understood how to multiply numbers. In such a framework making the test \emph{adaptive}, i.e., picking a next question on the basis of the current knowledge level of the test taker is also very natural. The expected information gain for the available questions might be naturally used to select the question leading to the more informative results about the skill level of the test taker. This might take place before the answer to the question on the basis of the expectations about the possible answers.

A critical point when coping with such approaches is to provide a realistic assessment for the probabilistic parameters associated with the modelling of the relations between the questions and the skills. Having to provide sharp numerical values for these probabilities might be difficult. As the skill is typically a latent quantity, complete data are not available for a statistical learning of the parameters and the elicitation should therefore demanded to experts (e.g., a teacher). Yet, it might be not obvious to express such a domain knowledge by single numbers and a more robust elicitation, such as a probability interval, might add realism and robustness to the modelling process \cite{hajek2012rationality}. With such generalized assessments of the parameters a Bayesian network simply becomes a \emph{credal} network \cite{antonucci2010d}. Yet, the counterpart of such increased realism is the higher computational complexity characterizing the inference in credal networks \cite{maua14jair}. This is especially the case when coping with information theoretic measures such a information gain, whose computation might lead to complex non-linear optimization tasks \cite{mangili2017b}. The goal of this paper is to investigate the potential of alternatives to the information-theoretic scores driving the question selection in adaptive tests based on directed graphical models, no matter whether these are Bayesian or credal networks. In particular, we consider a family of scores based on the (expected) mode of the posterior distributions over the skills. We show that, when coping with credal networks, the computation of these scores can be reduced to the simpler optimization tasks involving multi-linear optimizations for which powerful approximation schemes are already available \cite{antonucci2013a}. Moreover, we show that these scores benefit of better explainability properties, thus allowing for a more transparent process in the question selection.

\begin{figure}[htp!]
\centering
\begin{tikzpicture}[scale=1]
\node[] ()  at (-5,0.5) {\tiny $P(S=1|Q=1)=0.5$};
\node[] ()  at (-5,-2) {\tiny $P(S=1|Q=1)=0.5$};
\node[] ()  at (-5,-3) {\tiny $P(S=1|Q=1)=0.5$};
\node[] ()  at (3,0.5) {\tiny $P(S=1)=0.5$};
\node[] ()  at (0,-3) {\tiny $2x3=?$ \quad $2x3=?$\quad  $2x3=?$};
\node[] ()  at (0,-2) {\tiny $P(Q=1|S=1)=0.3$ \quad $0.4$\quad  $0.5$};
\node[] ()  at (0,-2.5) {\tiny $P(Q=1|S=0)=0.3$ \quad  $0.4$ \quad  $0.5$};
\node[nodo] (s)  at (0,0.5) {$S$};
\node[Qnodo] (q1)  at (-2.,-1) {$Q_1$};
\node[Qnodo] (q2)  at (0,-1) {$Q_2$};
\node[Qnodo] (q3)  at (2,-1) {$Q_3$};
\draw[arco] (s) -- (q1);
\draw[arco] (s) -- (q2);
\draw[arco] (s) -- (q3);
\end{tikzpicture}
\caption{Ad adaptive testing scheme to evaluate the multiplication (Boolean) skill $S$ by three (Boolean) questions. Probabilities are assessed by the Bayesian network on the right}
    \label{fig:minicat}
\end{figure}
*** espandere figura ***
\newpage 
The paper is organized as follows. A critical discussion about the existing work in this area is in Section \ref{sec:work}. The necessary background material is reviewed in Section \ref{sec:background}. The adaptive testing concepts are introduced in Section \ref{sec:cat} and specialized to graphical models in \ref{sec:bncat}. The technical part of the paper is in Section \ref{sec:mode}, where the new scores are discussed and specialized to the credal case, while the experiments are in Section \ref{sec:exp}. Conclusions and outlooks are in Section \ref{sec:conc}.

\section{Related Work}\label{sec:work}
Modelling a test as a process involving latent variables is a classical approach since the proposal of the classical \emph{item response theory}, that has been widely used even to implement adaptive sequences \cite{xxx}. Despite its success related to the ease of implementation and inference, IRT might be inadequate when coping with multiple latent skill, especially when these are dependent. For such cases, interpretable graphical models such as Bayesian networks have been resulted more effective [x]. The development of (computer) adaptive tests is a classical topic for AI \cite{xx}. Apart from graphical models, adaptive algorithms have been proposed within a genetic optimization framework. The idea of using Bayesian network in grading environment is not new. xxx. As already noticed in the previous section, the latent nature of the student skills might hinder the direct statistical learning of the parameters of such models. This naturally led some authors to consider missing data modelling. In particular the work of xxxx Vomlel related to EM algorithm in Bayesian networks is a notable example of this kind of approaches. Some other authors considered even more extreme directions involving missingness even about the answers such as \cite{bachrach2012grade}. This is also related to the notion of monotonicity \cite{plajner2020monotonicity}. The potential of these models in terms of explainability of the results has been emphasized in the work Work of Darwiche about same-probability problem. *** completare *** Bayesian networks (BNs) \cite{koller2009} have been used to model the knowledge driving such intelligent systems \cite{almond2015bayesian}. However, collecting large sets of reliable data in educational domains may be difficult and time consuming (e.g., a course with few students, or taught for the first time), and the quantification should be based on expert knowledge only. To elicit a Bayesian network, an expert might face questions like: ``\emph{which is the probability of a student with a particular knowledge level giving the right answer to a question?}''. Giving sharp probabilities for questions of this kind can be problematic for an expert, whose knowledge is mostly qualitative (e.g., ``\emph{a right answer is very unlikely}''). Fuzzy linguistic approaches represent a viable, non-numerical, way to address these issues \cite{badaracco2013fuzzy}. To stick within the probabilistic framework, verbal-numerical probability scales associated with sharp values \cite{renooij1999talking} or intervals \cite{walley1991statistical} have been also proposed. a generalization of BNs based on the imprecise probability theory \cite{walley1991statistical}, where local parameters are defined by set-valued probabilities. This simplifies the elicitation process and offers a more reliable handling of the related uncertainty. Moving from BNs to CNs implies two main issues: (i) numerical inferences will be interval-valued too, thus making debatable both the decision criterion \cite{troffaes} an the information measures \cite{klir1999uncertainty} to adopt; and (ii) inference tasks in CNs typically belongs to higher complexity classes than their Bayesian counterparts \cite{maua14jair}. Both these issues are addressed by defining a computationally feasible procedure based on CNs to be used for practical implementation of intelligent systems solely specified by expert knowledge. To the best of our knowledge this is the first attempt to perform e-testing with models of this kind.

We focus on the application of CNs to \emph{computer adaptive testing} (CAT), i.e., an approach to e-testing that adjusts the sequence and the number of questions to the ability level of the test taker. CATs have the potential to make the test an individualised experience that challenges and does not discourage the test takers, as most of the questions are near their ability levels. Building upon \emph{item response theory} \cite{hambleton1985item}, the common background underpinning CATs, graphical modeling (such as BNs and CNs) offers a powerful language for describing complex multivariate dependencies between skills and rich tasks. Several researchers have exploited the potential of BNs both in adaptive and non-adaptive educational assessment \cite{vomlel2004building,plajner2015}. These authors focus on applications for which data are available to learn the model parameters. We regard this point as a serious limitation, possibly hindering CATs adoption by many teachers and instructors. The results are promising: CAT based on CNs is effective in reducing the number of questions while maintaining a high accuracy in the evaluation and the approximations introduced do not compromise the procedure's effectiveness.

\section{Background on Bayesian and Credal Networks}\label{sec:background}
We denote variables by Latin uppercase letters, while using lowercase for their generic values, and calligraphic for the set of their possible values. Thus, $x \in \mathcal{X}$ is a possible value of $X$. Here only variables with a finite number of possible values are considered.\footnote{Adaptive approaches such as IRT \cite{xx} adopt continuous latent variables.}

\subsection{Bayesian Networks}
A probability mass function over $X$ is denoted as $P(X)$, while $P(x)$ is the probability assigned to state $x$. Given a function $f$ of $X$, its expectation with respect to $P(X)$ is $\mathbb{E}_P(f):=\sum_{x\in\mathcal{X}} P(x) \cdot f(x)$. The expectation of $-\log_k[P(X)]$ is called \emph{entropy} and denoted also as $H_P(X)$.\footnote{To cope with zero probabilities, we set $0 \cdot \log_k 0 = 0$.} We assume $k:=|\mathcal{X}|$ to have the maximum of the entropy, achieved for uniform PMFs, equal to one.

Given a joint PMF $P(X,Y)$, the marginal PMF $P(Y)$ is obtained by summing out the other variable, i.e., $P(y)=\sum_{x\in\mathcal{X}} P(x,y)$. Conditional PMFs such as $P(X|y)$ are similarly obtained by Bayes's rule, i.e., $P(x|y)=P(x,y)/P(y)$ provided that $P(y)>0$. Notation $P(X|Y):=\{P(X|y)\}_{y\in\mathcal{Y}}$ is used for such conditional probability table (CPT). The entropy of a conditional PMF is defined as in the unconditional case and denoted as $H_P(X|y)$. The expected entropy is a weighted average of conditional entropies, i.e., $H_P(X|Y):=\sum_{y\in\mathcal{Y}} H_P(X|y) P(y)$. If $P(x,y)=P(x) P(y)$ for each $x\in\mathcal{X}$ and $y\in\mathcal{Y}$, variables $X$ and $Y$ are stochastically independent. A conditional formulation can be also considered.

We assume the set of variables $\bm{X}:=(X_1,\ldots,X_n)$ to be in one-to-one correspondence with a directed acyclic graph $\mathcal{G}$. For each $X\in\bm{X}$, the parents of $X$ are called predecessor of $X$ according to $\mathcal{G}$ and denoted as $\mathrm{Pa}_X$. The graph $\mathcal{G}$ together with the collection of CPTs $\{P(X|\mathrm{Pa}_X)\}_{X\in\bm{X}}$ provides a Bayesian network (BN) specification \cite{koller2009}. Under the Markov condition, that states that every variable is conditionally independent of its non-descendants non-parents given its parents, a BN compactly defines a joint PMF $P(\bm{X})$ that factorizes as $P(\bm{x})=\prod_{X\in\bm{X}} P(x|\mathrm{pa}_X)$.

\subsection{Credal Sets and Credal Networks}
A set of PMFs over $X$ is denoted as $K(X)$ and called \emph{credal set} (CS). Expectations based on CSs are the bounds of the PMF expectations with respect to the CS. Thus $\underline{E}_K[f]:=\inf_{P(X)\in K(X)} \mathbb{E}_P[f]$ and similarly for the supremum $\mathbb{E}_K$. Expectations of events are in particular caller lower and upper probabilities and denotes as $\underline{P}$ and $\overline{P}$. Notation $K(X|y)$ is used for a set of conditional CSs, while $K(X|Y):=\{K(X|y)\}_{y\in\mathcal{Y}}$ is a credal CPT (CCPT).

Analogously to a BN, a credal network (CN) \cite{cozman2000}  is specified by graph $\mathcal{G}$ together with a family of CCPTs $\{K(X|\mathrm{Pa}_X)\}_{X\in\bm{X}}$. A CN defines a joint CS $K(\bm{X})$ obtained by the set of xxx.

\section{Testing Algorithms}\label{sec:cat}
A typical test aims at evaluating the knowledge level of a test taker $\sigma$ on the basis of her answers to a number of questions. Let $\bm{Q}$ denote a repository of questions available to the instructor. The order and the number of questions from $\bm{Q}$ asked to $\sigma$ might not be defined in advance. We call \emph{testing algorithm} (TA) a procedure taking care of the selection of the sequence of questions asked to the test taker, and to decide when the test stops. Algorithm \ref{alg:ta} depicts a general TA scheme, with $\bm{e}$ denoting the array of the answers collected from the test taker $\sigma$.

@@

Boolean function ${\tt Stopping}$ decides whether the test should end, this choice being possibly based on the previous answers in $\bm{e}$. Trivial stopping rules might be based on the number of questions asked to the test takes (${\tt Stopping}(\bm{e})=1$ if and only if $|\bm{e}|>n$) or on the number of correct answers provided that a maximum number of questions is not exceeded. Function ${\tt Pick}$ selects instead the question to be asked to the student from the repository $\bm{Q}$. A TA is called \emph{adaptive} when this function takes into account the previous answers $\bm{e}$. Trivial non-adaptive strategies might consist in randomly picking an element of $\bm{Q}$ or following a fixed order. Function ${\tt Answer}$ is simply collecting (or simulating) the answer of test taker $\sigma$ to a particular question $Q$. In our assumptions, this answer is not affected by the previous answers to other questions.\footnote{Generalized setups where the quality of the student answer is affected by the previous answers will be discussed at the end of the paper. This might include a \emph{fatigue} model negatively affecting the quality of the answers when many questions have been already answered as well as the presence of \emph{revealing} questions that might improve the quality of other answers.}

Finally, ${\tt Evaluate}$ is a function returning the overall judgement of the test (e.g., a numerical grade or a pass/fail Boolean) on the basis of all the answers collected after the test termination. Trivial examples of such functions are the percentage of correct answers or a Boolean that is true when a sufficient number of correct answers has been provided. Note also that in our assumptions the TA is \emph{exchangeable}, i.e., the stopping rule, the question finder and the evaluation function are invariant with respect to permutations in $\bm{e}$. In other words, the same next question, the same evaluation and the same stopping decision is produced for any two students, who provided the same list of answers in two different orders.

A typical TA is required to achieve a reliable evaluation of the student $\sigma$ after the answers $\bm{e}$. As each answer is individually assumed to improve such quality, asking all the questions, no matter in which  order because of the exchangeability assumption, is an obvious choice. Yet, this might be impractical (e.g., because of time limitations) or just provide an unnecessary burden to the test taker. The goal of a good TA is therefore to trade off the accuracy of the evaluation and the number of questions.\footnote{In some generalized setups, other elements such as a \emph{serendipity} in choice in order to avoid tedious sequences of questions might be also considered \cite{xx}.}

\section{Adaptive Testing in Bayesian Networks}\label{sec:bncat}
%Item response theory is a popular *** Something about IRT here ***. Most of these approaches are based an a unidimensional representation of the student knowledge level. Proposals to cope with multidimensional IRTs have been considered, but this results critical when coping with dependencies between them. For this reason probabilistic modelling can be considered. This is the topic of the next section.

The general TA setup in Algorithm \ref{alg:ta} can be easily specialized to BNs as follows.
First, we identify the profile $\sigma$ of the test saker with the actual states of a number of latent discrete variables, called \emph{skills}. Let $\bm{S}=\{S_i\}_{j=1}^n$ denote these skill variables, and $\bm{s}_\sigma$ the actual values associated with the test take. Skills are typically ordinal variables, whose states corresponds to increasing knowledge levels. Questions in $\bm{Q}$ are still described as manifest variables whose actual values are returned by the {\tt answer} function. This is achieved by a (possibly stochastic) function of the actual profile $\bm{s}_\sigma$. Thus reflect the taker perspective, while the teacher has clearly no access to $\bm{s}_\sigma$. As a remark, note that we might coarsen the set of possible values $\mathcal{Q}$ each $Q\in \bm{Q}$: for instance, a multiple choice question with three options might have a single right, the two other answers being indistinguishable from the evaluation point of view.\footnote{The case of \emph{abstention} to an answer and the consequent problem of modelling the incompleteness is a topic we do not consider here for the sake of conciseness. Yet, general approaches in the spirit of xxx could be naturally considered.}

A joint PMF over the skills $\bm{S}$ and the questions $\bm{Q}$ is supposed to be available. In particular we assume this to correspond to a BN whose graph has the questions as leaf nodes. Thus, for each $Q\in\bm{Q}$, $Pa_Q \subseteq \bm{S}$ and we call $Pa_Q$ the \emph{scope} of question $Q$. Note that this assumption about the graph is simply reflecting a statement about the conditional independence between (the answer to) a question and all the other skills and questions given scope of the question. This basically means that the answers to other questions are not directly affecting the answer to a particular question, and this naturally follows from the exchangeability assumption.\footnote{Moving to other setups would not really be critical because of the property of observed nodes in Bayesian and credal networks, see for instance \cite{antonucci2009} or \cite{debock,bolt}.}

>>>>

\usepackage{algorithm,algorithmic,amsfonts,bm,booktabs,tikz,pgf,pgfplots,amsmath,soul,url,graphicx,amsmath,booktabs,booktabs,nicefrac,bm,colortbl,tikz,pgfplots,enumerate,xargs,optidef}

@@

%\captionsetup{compatibility=false}

@@

\title{A New Score for Adaptive Tests\\ in Bayesian and Credal Networks}

@@

Francesca Mangili \and \\

@@

\institute{Istituto Dalle Molle di Studi sull'Intelligenza Artificiale, Lugano, Switzerland

@@

A test is \emph{adaptive} when its sequence and number of questions is dynamically tuned on the basis of the estimated skills of the taker. Graphical models, such as Bayesian networks, are used for adaptive tests as they allow to model the uncertainty about the questions and the skills in an explainable fashion, especially when coping with multiple skills. A better elicitation of the uncertainty in the question/skills relations can be achieved by interval probabilities. This turns the model into a \emph{credal} network, thus making more challenging the inferential complexity of the queries required to select questions. This is especially the case for the information theoretic quantities used as \emph{scores} to drive the adaptive mechanism. We present an alternative family of scores, based on the mode of the posterior probabilities, and hence easier to explain. This makes considerably simpler the evaluation in the credal case, without significantly affecting the quality of the adaptive process. Numerical tests on synthetic and real-world data are used to support this claim.
\keywords{computer adaptive tests \and information theory \and credal networks \and Bayesian networks \and index of qualitative variation}

@@

A test or an exam can be naturally intended as a measurement process, with the questions acting as sensors measuring the skills of the test taker in a particular discipline. Such measurement is typically imperfect with the skills modelled as latent variables whose actual values cannot be revealed in a perfectly reliable way. The role of the questions, whose answers are regarded instead as manifest variables, is to reduce the uncertainty about the latent skills. Following this perspective, probabilistic models are an obvious framework to describe tests. Consider for instance the example in Figure \ref{fig:minicat}, where a Bayesian network evaluates the probability that the test taker knows how to multiply integers. In such framework making the test \emph{adaptive}, i.e., picking a next question on the basis of the current knowledge level of the test taker is also very natural. The information gain for the available questions might be used to select the question leading to the more informative results (e.g., according to Table \ref{tab:minicat}, $Q_1$ is more informative than $Q_2$ no matter what the answer is). This might also be done before the answer on the basis of expectations over the possible alternatives.

A critical point when coping with such approaches is to provide a realistic assessment for the probabilistic parameters associated with the modelling of the relations between the questions and the skills. Having to provide sharp numerical values for these probabilities might be difficult. As the skill is a latent quantity, complete data are not available for a statistical learning and a direct elicitation should be typically demanded to experts (e.g., a teacher). Yet, it might be not obvious to express such a domain knowledge by single numbers and a more robust elicitation, such as a probability interval (e.g., $P(Q_1=1|S_1=1)\in[0.85,0.95]$), might add realism and robustness to the modelling process \cite{hajek2012rationality}. With such generalized assessments of the parameters a Bayesian network simply becomes a \emph{credal} network \cite{antonucci2010d}. The counterpart of such increased realism is the higher computational complexity characterizing inference in credal networks \cite{maua14jair}. This is an issue especially when coping with information theoretic measures such an information gain, whose computation in credal networks might lead to complex non-linear optimization tasks \cite{mangili2017b}. The goal of this paper is to investigate the potential of alternatives to the information-theoretic scores driving the question selection in adaptive tests based on directed graphical models, no matter whether these are Bayesian or credal networks. In particular, we consider a family of scores based on the (expected) mode of the posterior distributions over the skills. We show that, when coping with credal networks, the computation of these scores can be reduced to the simpler optimization tasks involving multi-linear optimizations for which powerful approximation schemes are already available \cite{antonucci2013a}. Moreover, we show that these scores benefit of better explainability properties, thus allowing for a more transparent process in the question selection.

\begin{figure}[htp!]
\centering
\begin{tikzpicture}[scale=1]
%\node[] ()  at (4,0.5) {\tiny $P(S=1)=0.5$};
\node[] ()  at (5.5,0.5) {$P(Q_1=1|S=1)=0.9$};
\node[] ()  at (5.5,0.) {$P(Q_1=1|S=0)=0.3$};
\node[] ()  at (5.5,-0.5) {$P(Q_2=1|S=1)=0.6$};
\node[] ()  at (5.5,-1.) {$P(Q_2=1|S=0)=0.4$};
\node[nodo] (s)  at (0,0.5) {Knows multiplication ($S$)};
\node[Qnodo] (q1)  at (-2.,-1) {$10\times 5?$ ($Q_1$)};
\node[Qnodo] (q2)  at (+2,-1) {$13\times 14?$($Q_2$)};
\draw[arco] (s) -- (q1);
\draw[arco] (s) -- (q2);
\end{tikzpicture}
\caption{A Bayesian network over Boolean variables modelling a simple test to evaluate integer multiplication skill with two questions.}
\label{fig:minicat}
\end{figure}
\begin{table}[htbp!]
\centering
\caption{Posterior probabilities of the skill after one or two questions in the test based on the Bayesian network in Figure \ref{fig:minicat}. A uniform prior over the skill is considered. Probabilities are regarded as grades and sorted from the lowest one. Bounds to a perturbation $\epsilon=\pm 0.05$ of all the model input parameters are also reported.}
\begin{tabular}{@{}ccrrr@{}}
\toprule
$Q_1$&$Q_2$&$P(S=1|q_1,q_2)$&
$\underline{P}(S=1|q_1,q_2)$&
$\overline{P}(S=1|q_1,q_2)$
\\
\midrule
$0$&$0$&$0.087$&$0.028$&$0.187$\\
$0$&$-$&$0.125$&$0.052$&$0.220$\\
$0$&$1$&$0.176$&$0.092$&$0.256$\\
$-$&$0$&$0.400$&$0.306$&$0.506$\\
$-$&$1$&$0.600$&$0.599$&$0.603$\\
$1$&$0$&$0.667$&$0.626$&$0.708$\\
$1$&$-$&$0.750$&$0.748$&$0.757$\\
$1$&$1$&$0.818$& $0.784$ &$0.852$\\
\bottomrule
\end{tabular}
\label{tab:minicat}
\end{table}
The paper is organized as follows. A critical discussion about the existing work in this area is in Section \ref{sec:work}. The necessary background material is reviewed in Section \ref{sec:background}. The adaptive testing concepts are introduced in Section \ref{sec:cat} and specialized to graphical models in \ref{sec:bncat}. The technical part of the paper is in Section \ref{sec:mode}, where the new scores are discussed and specialized to the credal case, while the experiments are in Section \ref{sec:exp}. Conclusions and outlooks are in Section \ref{sec:conc}.

\section{Related Work}\label{sec:work}
Modelling a test as a process relating latent and manifest variables since the classical \emph{item response theory}, that has been widely used even to implement adaptive sequences \cite{embretson2013item}. Despite its success related to the ease of implementation and inference, IRT might be inadequate when coping with multiple latent skill, especially when these are dependent. This led researchers towards the area of probabilistic graphical models as tools to implement IRT in more complex setups \cite{mislevy1997graphical}. Eventually, Bayesian networks have been identified as a natural framework to models tests, even behind the IRT framework \cite{vomlel2004bayesian}, this being especially the case for adaptive models \cite{vomlel2004building} and coached solving \cite{conati1997line}. In order to cope with latent skills, some authors explored the application of EM-like approaches to these models \cite{plajner2020monotonicity}, this also involving the extreme situation of no ground truth information about the answers \cite{bachrach2012grade}.
As an alternative approach to the same issue, some authors considered relaxations of the Bayesian formalism, such as fuzzy models \cite{badaracco2013fuzzy} and imprecise probabilities \cite{mangili2017b}. This latter direction is the same direction we consider, but trying to overcome the computational limitations of that approach when coping with information-theoretic scores. Compared to the approach in \cite{chen2015computer}, that is focused in the Bayesian case only, we are interested in measures that could easily extended to the imprecise probability framework, this not being the case of the same-decision problem considered in that work.

\section{Background on Bayesian and Credal Networks}\label{sec:background}
We denote variables by Latin uppercase letters, while using lowercase for their generic values, and calligraphic for the set of their possible values. Thus, $x \in \mathcal{X}$ is a possible value of $X$. Here we only consider discrete variables.\footnote{IRT uses instead with continuous skills. Yet, when coping probabilistic models, having discrete skill does no prevent evaluations to range over a continuous domain. E.g., see Table \ref{tab:minicat}, where the grade corresponds to a (continuous) probability.}

\subsection{Bayesian Networks}
A probability mass function (PMF) over $V$ is denoted as $P(V)$, while $P(v)$ is the probability assigned to state $v$. Given a function $f$ of $V$, its expectation with respect to $P(V)$ is $\mathbb{E}_P(f):=\sum_{v\in\mathcal{V}} P(v) \cdot f(v)$. The expectation of $-\log_b[P(V)]$ is called \emph{entropy} and denoted also as $H(X)$.\footnote{We set $0 \cdot \log_b 0 = 0$ to cope with zero probabilities.} We set $b:=|\mathcal{V}|$ to have the maximum of the entropy, achieved for uniform PMFs, equal to one.

Given joint PMF $P(U,V)$, the marginal PMF $P(V)$ is obtained by summing out the other variable, i.e., $P(v)=\sum_{u\in\mathcal{U}} P(u,v)$. Conditional PMFs such as $P(U|v)$ are similarly obtained by Bayes's rule, i.e., $P(u|v)=P(u,v)/P(v)$ provided that $P(v)>0$. Notation $P(U|V):=\{P(U|v)\}_{v\in\mathcal{V}}$ is used for such conditional probability table (CPT). The entropy of a conditional PMF is defined as in the unconditional case and denoted as $H(U|v)$. The conditional entropy is a weighted average of entropies of the conditional PMFs, i.e., $H(U|V):=\sum_{v\in\mathcal{V}} H(U|v) P(v)$. If $P(u,v)=P(u) P(v)$ for each $u\in\mathcal{U}$ and $v\in\mathcal{V}$, variables $U$ and $V$ are independent. Conditional formulations are also considered.

We assume the set of variables $\bm{V}:=(V_1,\ldots,V_r)$ to be in one-to-one correspondence with a directed acyclic graph $\mathcal{G}$. For each $V\in\bm{V}$, the parents of $V$, i.e., the predecessors of $V$ in  $\mathcal{G}$, are denoted as $\mathrm{Pa}_V$. Graph $\mathcal{G}$ together with the collection of CPTs $\{P(V|\mathrm{Pa}_V)\}_{V\in\bm{V}}$ provides a Bayesian network (BN) specification \cite{koller2009}. Under the Markov condition, that states that every variable is conditionally independent of its non-descendants non-parents given its parents, a BN compactly defines a joint PMF $P(\bm{V})$ that factorizes as $P(\bm{v})=\prod_{V\in\bm{V}} P(v|\mathrm{pa}_V)$. Inference intended as the computation of the posterior PMF of a single (queried) variables given some evidence about other variables is in general NP-hard, but exact and approximate schemes are available (see \cite{koller2009} for details).

\subsection{Credal Sets and Credal Networks}
A set of PMFs over $V$ is denoted as $K(V)$ and called \emph{credal set} (CS). Expectations based on CSs are the bounds of the PMF expectations with respect to the CS. Thus $\underline{\mathbb{E}}[f]:=\inf_{P(V)\in K(V)} \mathbb{E}[f]$ and similarly for the supremum $\overline{\mathbb{E}}$. Expectations of events are in particular caller lower and upper probabilities and denoted as $\underline{P}$ and $\overline{P}$. Notation $K(U|v)$ is used for a set of conditional CSs, while $K(U|V):=\{K(U|v)\}_{v\in\mathcal{V}}$ is a credal CPT (CCPT).

Analogously to a BN, a credal network (CN) is specified by graph $\mathcal{G}$ together with a family of CCPTs $\{K(V|\mathrm{Pa}_V)\}_{V\in\bm{V}}$ \cite{cozman2000}. A CN defines a joint CS $K(\bm{V})$ corresponding to all the joint PMFs induced by BNs whose CPTs are consistent with the CN CCPTs. For CNs, we intend inference as the computation of the lower and upper posterior probabilities. The task generalizes BN inference being therefore NP-hard, see  \cite{maua2014jair} for a deeper characterization.

\section{Testing Algorithms}\label{sec:cat}
A typical test aims at evaluating the knowledge level of a test taker $\sigma$ on the basis of her answers to a number of questions. Let $\bm{Q}$ denote a repository of questions available to the instructor. The order and the number of questions picked from $\bm{Q}$ to be asked to $\sigma$ might not be defined in advance. We call \emph{testing algorithm} (TA) a procedure taking care of the selection of the sequence of questions asked to the test taker, and to decide when the test stops. Algorithm \ref{alg:ta} depicts a general TA scheme, with $\bm{e}$ denoting the array of the answers collected from test taker $\sigma$.

@@

Boolean function ${\tt Stopping}$ decides whether the test should end, this choice being possibly based on the previous answers in $\bm{e}$. Trivial stopping rules might be based on the number of questions asked to the test takes (${\tt Stopping}(\bm{e})=1$ if and only if $|\bm{e}|>n$) or on the number of correct answers provided that a maximum number of questions is not exceeded. Function ${\tt Pick}$ selects instead the question to be asked to the student from the repository $\bm{Q}$. A TA is called \emph{adaptive} when this function takes into account the previous answers $\bm{e}$. Trivial non-adaptive strategies might consist in randomly picking an element of $\bm{Q}$ or following a fixed order. Function ${\tt Answer}$ is simply collecting (or simulating) the answer of test taker $\sigma$ to a particular question $Q$. In our assumptions, this answer is not affected by the previous answers to other questions.\footnote{Generalized setups where the quality of the student answer is affected by the previous answers will be discussed at the end of the paper. This might include a \emph{fatigue} model negatively affecting the quality of the answers when many questions have been already answered as well as the presence of \emph{revealing} questions that might improve the quality of other answers \cite{laitusis2007examination}.}

Finally, ${\tt Evaluate}$ is a function returning the overall judgement of the test (e.g., a numerical grade or a pass/fail Boolean) on the basis of all the answers collected after the test termination. Trivial examples of such functions are the percentage of correct answers or a Boolean that is true when a sufficient number of correct answers has been provided. Note also that in our assumptions the TA is \emph{exchangeable}, i.e., the stopping rule, the question finder and the evaluation function are invariant with respect to permutations in $\bm{e}$ \cite{sawatzky2016accuracy}. In other words, the same next question, the same evaluation and the same stopping decision is produced for any two students, who provided the same list of answers in two different orders.

A typical TA is required to achieve a reliable evaluation of the student $\sigma$ after the answers $\bm{e}$. As each answer is individually assumed to improve such quality, asking all the questions, no matter in which  order because of the exchangeability assumption, is an obvious choice. Yet, this might be impractical (e.g., because of time limitations) or just provide an unnecessary burden to the test taker. The goal of a good TA is therefore to trade off the accuracy of the evaluation and the number of questions.\footnote{In some generalized setups, other elements such as a \emph{serendipity} in choice in order to avoid tedious sequences of questions might be also considered \cite{badran2019adaptive}.}

\section{Adaptive Testing in Bayesian and Credal Networks}\label{sec:bncat}
The general TA setup in Algorithm \ref{alg:ta} can be easily specialized to BNs as follows. First, we identify the profile $\sigma$ of the test taker with the actual states of a number of latent discrete variables, called \emph{skills}. Let $\bm{S}=\{S_i\}_{j=1}^n$ denote these skill variables, and $\bm{s}_\sigma$ the actual values associated with the test take. Skills are typically ordinal variables, whose states corresponds to increasing knowledge levels. Questions in $\bm{Q}$ are still described as manifest variables whose actual values are returned by the {\tt answer} function. This is achieved by a (possibly stochastic) function of the actual profile $\bm{s}_\sigma$. Thus reflect the taker perspective, while the teacher has clearly no access to $\bm{s}_\sigma$. As a remark, note that we might coarsen the set of possible values $\mathcal{Q}$ each $Q\in \bm{Q}$: for instance, a multiple choice question with three options might have a single right, the two other answers being indistinguishable from the evaluation point of view.\footnote{The case of \emph{abstention} to an answer and the consequent problem of modelling the incompleteness is a topic we do not consider here for the sake of conciseness. Yet, general approaches in the spirit of xxx could be naturally considered.}

A joint PMF over the skills $\bm{S}$ and the questions $\bm{Q}$ is supposed to be available. In particular we assume this to correspond to a BN whose graph has the questions as leaf nodes. Thus, for each $Q\in\bm{Q}$, $Pa_Q \subseteq \bm{S}$ and we call $Pa_Q$ the \emph{scope} of question $Q$. Note that this assumption about the graph is simply reflecting a statement about the conditional independence between (the answer to) a question and all the other skills and questions given scope of the question. This basically means that the answers to other questions are not directly affecting the answer to a particular question, and this naturally follows from the exchangeability assumption.\footnote{Moving to other setups would not really be critical because of the property of observed nodes in Bayesian and credal networks, see for instance \cite{antonucci2009,bolt}.}

As the available data are typically incomplete because of the latent nature of the skills, dedicated learning strategies, such as various form of constrained EM should be considered to train a BN from data. We refer the reader to the various contributions of Plajner and Vomlel in this field (e.g., \cite{plajner2020monotonicity}) for a complete discussion of that approach. Here we assume the BN quantification already available.

In such a BN framework, ${\tt Stopping}(\bm{e})$ might be naturally based on an evaluation of the posterior PMF $P(\bm{S}|\bm{Q}=\bm{e})$, this being also the case for ${\tt Evaluate}$. Regarding the question selection, ${\tt Pick}$ might be similarly based on the (posterior) CPT $P(\bm{S}|Q,\bm{Q}=\bm{e})$, whose values for the different answers to $Q$ might be weighted by the marginal $P(Q|\bm{e})$. More specifically, entropies and conditional entropies are considered by Algorithm \ref{alg:bnta}, while the evaluation is based on a conditional expectation for a given utility function.
